---
title: "Lab 4"
subtitle: "Permutation and bootstrapping in R"
toc: true
editor: source
format:
  html:
    other-links:
      - text: Download Source
        href: lab4.qmd
      - text: Eel Grass Restoration Data
        href: data/eelgrass.csv
editor_options: 
  chunk_output_type: console
---

## Learning objectives

In today's lab you will...

1.  Run a **permutation test** to answer a "yes/no?" question
2.  **Bootstrap a confidence interval** to answer a "how much?" question

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
theme_set(theme_classic(14))
set.seed(123)

```

## Eel grass restoration

Recall the eel grass restoration example from yesterday's lecture. The data recorded whether attempts to restore eel grass were successful based on the method used (garden staple or popsicle stick) [@fig-eelgrass-data].

```{r, results='hide'}
#| label: read-eelgrass-data

restoration <- read_csv("week4/data/eelgrass.csv",
                        show_col_types = FALSE)
glimpse(restoration)

```

```{r, results='hide'}
#| label: fig-eelgrass-data
#| fig-cap: "Success of eelgrass restoration plots by restoration method."

ggplot(restoration, aes(treatment, fill = success_fct)) +
  geom_bar() +
  scale_fill_manual(values = c("firebrick", "cornflowerblue")) +
  labs(x = "Restoration method",
       y = "Sites",
       fill = "Outcome") +
  theme(legend.position = "inside",
        legend.position.inside = c(1, 1),
        legend.justification = c(1, 1))

```

## Permutation test

We want to know if the popsicle stick method works better for restoration than garden staples. This is a *"yes/no?"* question, so we'll use a permutation test for our hypothesis. Recall the steps for hypothesis testing:

1.  Identify the TEST STATISTIC
2.  State your NULL and ALTERNATIVE hypotheses
3.  Calculate the OBSERVED test statistic
4.  Estimate the NULL DISTRIBUTION
5.  Calculate P-VALUE
6.  Compare p-value to CRITICAL THRESHOLD

### Identify the test statistic

**Q1: What is the appropriate test statistic for this question?**

Difference in proportions

### State your null and alternative hypotheses

**Q2: What are your null and alternative hypotheses?**

H0: There is no difference in restoration success between the garden staple and Popsicle stick method.

HA: The Popsicle stick method is associated with higher restoration success compared to the garden staple method.

### Calculate the observed test statistic

**Q3: How would you calculate the test statistic *for the sample*?**

diff_prop

```{r}
#| label: q3

ts_sample<- restoration %>% 
  group_by(treatment) %>% 
  summarise(test_stat = sum(success)/sum(n()))

diff_prop <- ts_sample[[2, 2]] - ts_sample[[1, 2]]

```

### Estimate the null distribution

This is the key part of a permutation test! Remember, our goal is to estimate the distribution of possible outcomes *under the null hypothesis*. To do that, we have to break the association between treatment and outcome.

**Q4: What column should we shuffle to break the association between treatment and outcome?**

We should shuffle treatment column

**Q5: Fill in the following code to perform one permutation and calculate the test statistic.**

```{r}
#| label: q5
#| eval: false

one_permutation <- restoration %>% 
  mutate(new_treatment = sample(treatment, 
                      size = length(restoration$treatment), 
                      replace = FALSE))

permutation_props <- one_permutation %>% 
  group_by(new_treatment) %>% 
  summarise(test_stat = sum(success)/sum(n()))

permutation_diff_props <- permutation_props$test_stat[2] - permutation_props$test_stat[1]

permutation_diff_props

```

That gives us the value of the test statistic for just one permutation. To get a distribution, we have to repeat the process many times. Let's do it 1,000 times.

**Q6: Fill in the following code to perform 1,000 permutations and estimate the null distribution**

```{r}
#| label: q6
#| eval: false

permute <- function(i) {
  one_permutation <- restoration %>% 
  mutate(new_treatment = sample(treatment, 
                      size = length(restoration$treatment), 
                      replace = FALSE))

permutation_props <- one_permutation %>% 
  group_by(new_treatment) %>% 
  summarise(test_stat = sum(success)/sum(n()))

permutation_diff_props <- permutation_props$test_stat[2] - permutation_props$test_stat[1]

permutation_diff_props
}

null_distribution <- map_dbl(1:1000, permute)

```

**Q7: Visualize the null distribution using a histogram and show where the observed test statistic falls**

```{r}
#| label: q7

null_distribution_df <- data.frame(null_distribution)

null_hist <- ggplot(null_distribution_df, aes(x=null_distribution)) + 
  geom_histogram() + 
  geom_vline(xintercept = diff_prop, linetype = "dashed", color = "red")

null_hist

```

### Calculate the p-value

The p-value is *the probability of a test statistic [at least]{.underline} as extreme as the observed, given the null hypothesis*. In other words, what proportion of the null distribution exceeds the observed?

**Q8: Calculate the p-value using the null distribution and observed test statistic**

```{r}
#| label: q8
#| eval: false

num_greater <- sum(null_distribution_df$null_distribution > diff_prop) # When estimating p value choose numbers greater than diff of prop

pvalue <- num_greater/sum(length(null_distribution_df$null_distribution))

pvalue
```

### Interpret the p-value

When interpreting the p-value, we compare it to a *critical threshold*, usually denoted with $\alpha$. By convention, we usually set $\alpha$ to 0.05.

**Q10:** **Given** $p \gt \alpha$**,** **which of the following statements is a correct interpretation and why?**

-   Our evidence is consistent with the hypothesis that restoration method does not influence restoration outcome

-   We cannot reject the hypothesis that restoration method does not influence restoration outcome

-   This is the most accurate response if my p value is greater than the threshold

-   This essentially means that I do not have enough evidence to reject the possibility of there being not difference between both treatments. I fail to reject the null hypothesis.

## Bootstrap confidence interval

Now let's answer a "how much?" question. We want to estimate an interval that we think contains the population parameter. For that, we use bootstrapping.

Recall the steps for bootstrapping:

1.  Identify the TEST STATISTIC
2.  Substitute sample for population and draw BOOTSTRAP SAMPLES
3.  Estimate the BOOTSTRAP DISTRIBUTION of the test statistic
4.  Calculate the CONFIDENCE INTERVAL

### Identify the test statistic

**Q11: What is the appropriate test statistic for this question?** - The difference of proportions

### Draw bootstrap samples

This is the key part of bootstrapping! Remember, our goal is to estimate the variation of our population's parameter due to sampling. To do that, we simulate the process of re-doing our experiment, using the original sample as a substitute for the population. To "re-do" our experiment, we have to keep the association between treatment and outcome.

**Q12: Fill in the following code to draw one bootstrap sample.**

```{r}
#| label: q12
#| eval: false

one_bootstrap <- restoration %>% 
  group_by(treatment) %>% 
  mutate( new_success = sample(success,
                      size = length(success),
                      replace = TRUE)) %>% 
  ungroup()

```

**Q13: Fill in the following code to draw 1,000 bootstrap samples.**

::: callout-tip
`list_rbind()` will take a list of data frames and bind them row-wise into one data frame.
:::

```{r}
#| label: q13
#| eval: false

bootstrap <- function(i) {
  one_bootstrap <- restoration %>% 
  group_by(treatment) %>% 
    mutate(new_success = sample(success,
                                size = length(success),
                                replace = TRUE)) %>% 
  ungroup() %>% 
    mutate(trial = i)
}

bootstrap_samples <- map(1:1000, bootstrap) %>% 
  list_rbind()

```

### Estimate the bootstrap distribution of the test statistic

**Q14: Fill in the following code to calculate the test statistic for each bootstrap sample.**

```{r}
#| label: q14

test_stat_per_trial <- bootstrap_samples %>% 
  group_by(trial, treatment) %>% 
  summarise(prop_success = mean(new_success), .groups = "drop_last") %>% 
  summarise(diff_prop = prop_success[2] - prop_success[1]) 

```

**Q15: Visualize the bootstrapped distribution of the test statistic.**

```{r}
#| label: q15

boot_hist <- ggplot(test_stat_per_trial, aes(x=diff_prop)) + 
  geom_histogram() 

boot_hist


```

### Calculate the confidence interval

A confidence interval (CI) is a range we are confident contains the population parameter. The bootstrapped distribution of the test statistic describes where we expect the population parameter to fall. So a 95% confidence interval, for example, spans the range from the 2.5% quantile of the bootstrap distribution to the 97.5% quantile.

**Q16: Find the bounds of the 95% CI.**

::: callout-tip
The `quantile()` function finds quantiles. It's vectorized over the parameter `probs`, so you can find multiple quantiles at once
:::

```{r}
#| label: q16
#| eval: false

restoration_ci <- quantile(test_stat_per_trial$diff_prop, c(0.025, 0.975))
restoration_ci

```

**Q17: Update your visual from Q15 to include the observed test statistic with a solid line and the confidence interval represented with dotted lines.**

```{r}
#| label: q17
boot_hist <- ggplot(test_stat_per_trial, aes(x=diff_prop)) + 
  geom_histogram() +
  geom_vline(xintercept = c(restoration_ci[1], restoration_ci[2]), linetype = "dashed", color = "red")
  
boot_hist

```

## Permutation vs bootstrap

The visualization you created for Q7 shows the null distribution of the test statistic. The visualization you created for Q17 shows the bootstrapped distribution of the test statistic.

**Q18: What did you do to make the null distribution center on zero? Specifically, what code?**

This occured when I broke the association between treatment and outcome. And then sampled without replacement.

**Q19: What did you do to make the bootstrap distribution center on the observed test statistic? Specifically, what code?**

This occurred when I preserved the association between treatment and outcome by re sampling success within each treatment. This caused the difference of proportions outcome to consistently be based on the initial proportions in each treatment, therefore bootstrap samples deviate from that initial value and center around it. I sampled with replacement. And I preserved the relationship between treatment and outcome.

**Q20: What would happen to your bootstrap distribution if you sampled without replacement?**

If I sampled without replacement the proportions of success and failure would remain exactly the same. I would only have the initial pool of success and failure values to assign to each treatment leading to the same proportions for every bootstrap sample.
